---
title: "Data 583 Life Expectancy (WHO)"
author: "Justin Chan, Kenny Tong, Viji Rajagopalan"
date: "7 Mar, 2023"
output: pdf_document
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EDA


### Original Dataset Summary & Initial Data Screening
Purpose : Let's take a snapshot of the original dataset and have a rough idea of its record
```{r}
le <- read.csv("dataset/LifeExpectancy.csv")
summary(le)
```
Let's look at the dataset dimension first
```{r}
dim(le)
```

Then, have a quick overall screening of the dataset
```{r}
head(le,5)

```

Here is another view :
```{r}
str(le)
```

From the above broad view, the following Conclusion/Key Findings are reached : 

- The records range is from Year 2000 to 2015
- Columns with NA : Life Expectancy, Adult Mortality, Alcohol, Hep B, BMI, Polio, Total exp, Dip, GDP, Population, thinness..1.19, thinness.5.9, Income.composition.of.resources, Schooling
- 'Status' Column is of the "character" data type, with values "Developing" and "Developed".  We will introduce a new column 'Status.val' to be the factor value of 'Status' for better analysis..
- 'Percentage Expenditure' has a mean value of `r mean(le$percentage.expenditure)` and max. value of `r max(le$percentage.expenditure)`.  Spending on health is more than the GDP per capita? Look into the column definition : Expenditure on health as a percentage of Gross Domestic Product per capita(%). The data of such magnitude simply does not quite make sense. Cross check with other references (e.g. the World Bank https://data.worldbank.org/indicator/SH.XPD.CHEX.GD.ZS).  OK, let's conclude that we have hesitation about the reliability/interpretation of the value of this column, and probably would drop and skip this column for the rest of this analysis.
- 'Population' and 'GDP' have a relatively large scale, compared with all other columns.  So, we may need to scale these two columns.

Now, let's do some data wrangling based on the above conclusions :
```{r}
# Create a new column Status.val to represent the Status column with number
le$Status.val <- ifelse(le$Status == "Developed",1,0)

# Create a new column as the scaled version of the GDP & Population, 
le$GDP_scaled = scale(le$GDP)
le$Population_scaled = scale(le$Population)

# Remove the unreliable column 
le <- subset(le,select=-c(percentage.expenditure))

```

### Null Value Analysis and Handling
Purpose : Investigate the and determine how to handle the null value in the data set

There are `r dim(le)[1]` no. of rows in the dataset.  Let's set the threshold of 20% as the max. proportion of null column to be allowed in a data column.  That means, columns with na over 20% will be dropped.  The threshold is then  `r dim(le)[1] * 0.2`.  So, the following 'Population' column will be dropped.



```{r}
#head(le)
le <- subset(le,select=-c(Population))
# also Population_Scaled
le <- subset(le,select=-c(Population_scaled))

```


For the other values, we will set the na to the respective column mean for the subsequent analysis.

```{r message=FALSE, warning=FALSE}

for(i in 1:ncol(le)) {                                   # Replace NA in all columns
  le[ , i][is.na(le[ , i])] <- mean(le[ , i], na.rm = TRUE)
}

```

Now, all na have been handled!  Let's continue our analysis.


### Overall General Life Expectancy Trend

Purpose : Do some visualisation to explore and identify the general data pattern, trends and clusters, etc

```{r fig.height=3, fig.width=3, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)

le %>%
  group_by(Year) %>%
  summarise(Life.expectancy = mean(Life.expectancy)) %>%
  ggplot(aes(x=Year,
               y=Life.expectancy)) +    
  geom_line()

```

Findings :

- The general life expectancy has been steadily increasing duration the year
- Average Life expectancy increase from about 67 to 71.5 in 15 years.



```{r fig.height=3, fig.width=3, message=FALSE, warning=FALSE}
le %>%
  group_by(Status) %>%
  summarise(Life.expectancy = mean(Life.expectancy)) %>%
  ggplot(aes(x=Status,
               y=Life.expectancy)) +    
  geom_bar(stat = "identity")

```
Finding :

- Life expectancy of Developed countries are significantly higher than that of Developing countries.

```{r fig.height=3, message=FALSE, warning=FALSE}
le.pivot <- pivot_longer(le,c(Adult.Mortality,under.five.deaths,infant.deaths),names_to='Mortality.Group',values_to='Mortality.Rate')
require(gridExtra)

le.pivot.area <- le.pivot %>%
  group_by(Year,Mortality.Group) %>%
  summarise(Mortality.Rate = mean(Mortality.Rate)) %>%
  ggplot(aes(x=Year,
             y=Mortality.Rate,
             fill=Mortality.Group)) +
  geom_area(position="stack",stat="identity")

le.pivot.line <- le.pivot %>%
  group_by(Year,Mortality.Group) %>%
  summarise(Mortality.Rate = mean(Mortality.Rate)) %>%
  ggplot(aes(x=Year,
             y=Mortality.Rate,
             color=Mortality.Group)) +
  geom_line()

grid.arrange(le.pivot.area, le.pivot.line, ncol=2)
```

Findings :

- The mortality rate of all three age groups are generally decreasing as a whole
- The mortality rate of the adult group, however, have fluctuation within the period



